


---
title: "Salmon Analysis - ZM"
output: html_document
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = TRUE)
    library(DESeq2)
    library("apeglm")
    library("ashr")
    library(org.Hs.eg.db)

    load("../../data/ZM_data.RData")
    dds
    results <- results(dds)
    resultsNames(dds)

    add_annotations_to_results <- function(res) {
    ens.str <- substr(rownames(res), 1, 15)
    res$symbol <- mapIds(org.Hs.eg.db,
                        keys=ens.str,
                        column="SYMBOL",
                        keytype="ENSEMBL",
                        multiVals="first")
    res$entrez <- mapIds(org.Hs.eg.db,
                        keys=ens.str,
                        column="ENTREZID",
                        keytype="ENSEMBL",
                        multiVals="first")
    return (res)
    }
```
        
# ZM Analysis {.tabset}

## TIME - ZM T0 to T16

### Unrestricted
#### Summary
```{r}
results_t16_t0 <- results(dds, contrast = c("timepoint", "t16", "t0"))

results_t16_t0 = add_annotations_to_results(results_t16_t0)

summary(results_t16_t0)

```

### Restricted
#### Summary
```{r}
results_t16_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t16", "t0")
)
summary(results_t16_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t16_t0_restricted)
```

        

### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t16_t0_shrunk <- lfcShrink(dds, coef="timepoint_t16_vs_t0", type="apeglm")
summary(results_t16_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t16_t0_shrunk)
```

        

#### ashr
```{r}
results_t16_t0_shrunk <- lfcShrink(dds, coef="timepoint_t16_vs_t0", type="ashr")
summary(results_t16_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t16_t0_shrunk)
```

        


### P-Adjusted Analysis
```{r}
## results_t0_to_t48
resSig <- subset(results_t16_t0, padj < 0.1) 
```

#### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

#### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```

## TIME - ZM T0 to T24

### Unrestricted
#### Summary
```{r}
results_t24_t0 <- results(dds, contrast = c("timepoint", "t24", "t0"))

results_t24_t0 = add_annotations_to_results(results_t24_t0)

summary(results_t24_t0)

```

### Restricted
#### Summary
```{r}
results_t24_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t24", "t0")
)
summary(results_t24_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t24_t0_restricted)
```

        

### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t24_t0_shrunk <- lfcShrink(dds, coef="timepoint_t24_vs_t0", type="apeglm")
summary(results_t24_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t24_t0_shrunk)
```

        

#### ashr
```{r}
results_t24_t0_shrunk <- lfcShrink(dds, coef="timepoint_t24_vs_t0", type="ashr")
summary(results_t24_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t24_t0_shrunk)
```

        


### P-Adjusted Analysis
```{r}
## results_t0_to_t48
resSig <- subset(results_t24_t0, padj < 0.1) 
```

#### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

#### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        


## TIME - ZM T0 to T36

### Unrestricted
#### Summary
```{r}
results_t36_t0 <- results(dds, contrast = c("timepoint", "t36", "t0"))

results_t36_t0 = add_annotations_to_results(results_t36_t0)

summary(results_t36_t0)

```

### Restricted
#### Summary
```{r}
results_t36_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t36", "t0")
)
summary(results_t36_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t36_t0_restricted)
```

        

### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t36_t0_shrunk <- lfcShrink(dds, coef="timepoint_t36_vs_t0", type="apeglm")
summary(results_t36_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t36_t0_shrunk)
```

        

#### ashr
```{r}
results_t36_t0_shrunk <- lfcShrink(dds, coef="timepoint_t36_vs_t0", type="ashr")
summary(results_t36_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t36_t0_shrunk)
```

        


### P-Adjusted Analysis
```{r}
## results_t0_to_t48
resSig <- subset(results_t36_t0, padj < 0.1) 
```

#### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

#### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        


## TIME - ZM T0 to T48

### Unrestricted
#### Summary
```{r}
results_t48_t0 <- results(dds, contrast = c("timepoint", "t48", "t0"))

results_t48_t0 = add_annotations_to_results(results_t48_t0)

summary(results_t48_t0)

```

```{r}
results_t48_t0_p0.5 <- results(dds, contrast = c("timepoint", "t48", "t0"), alpha=0.05)
summary(results_t48_t0_p0.5)
```

### Restricted
#### Summary
```{r}
results_t48_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t48", "t0")
)
summary(results_t48_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t48_t0_restricted)
```

        

### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t48_t0_shrunk <- lfcShrink(dds, coef="timepoint_t48_vs_t0", type="apeglm")
summary(results_t48_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t48_t0_shrunk)
```

        

#### ashr
```{r}
results_t48_t0_shrunk <- lfcShrink(dds, coef="timepoint_t48_vs_t0", type="ashr")
summary(results_t48_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t48_t0_shrunk)
with(results_t48_t0_shrunk[!is.na(results_t48_t0_shrunk$padj) & results_t48_t0_shrunk$padj<0.05, ], points(baseMean, log2FoldChange, col="red", pch=20))

```

        


#### P-Adjusted Analysis
```{r}
## results_t0_to_t48
resSig <- subset(results_t48_t0, padj < 0.05) 
summary(resSig)
```

##### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

##### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        

        