

---
title: "Salmon Analysis - ZM_r1to6 "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library("apeglm")
library("ashr")
library(EnhancedVolcano)

library(org.Hs.eg.db)

load("../../data/ZM_r1to6_data.RData")
dds
results <- results(dds)
resultsNames(dds)

add_annotations_to_results <- function(res) {
ens.str <- substr(rownames(res), 1, 15)
res$symbol <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="SYMBOL",
                    keytype="ENSEMBL",
                    multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="ENTREZID",
                    keytype="ENSEMBL",
                    multiVals="first")
return (res)
}
```
        
# ZM Analysis {.tabset}



## Time -- ZM T0 to T16

### Unrestricted
#### Summary
```{r}
results_t16_t0 <- results(dds, contrast = c("timepoint", "t16", "t0"))

results_t16_t0 = add_annotations_to_results(results_t16_t0)

summary(results_t16_t0)

```

### Restricted
#### Summary
```{r}
results_t16_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t16", "t0")
)
summary(results_t16_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t16_t0_restricted)
```

        

#### V- Plot

```{r}
results_t16_t0_restricted <- add_annotations_to_results(results_t16_t0_restricted)
results_t16_t0_restricted$adjp_neglog10 <- -log10(results_t16_t0_restricted$padj)
selected_genes <- as.character(results_t16_t0_restricted$symbol)

EnhancedVolcano(results_t16_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'adjp_neglog10',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t16_t0_shrunk <- lfcShrink(dds, coef="timepoint_t16_vs_t0", type="apeglm")

summary(results_t16_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t16_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t16_t0_shrunk <- add_annotations_to_results(results_t16_t0_shrunk)
selected_genes <- as.character(results_t16_t0_shrunk$symbol)
results_t16_t0_shrunk$adjp_neglog10 <- -log10(results_t16_t0_shrunk$pvalue)

EnhancedVolcano(results_t16_t0_shrunk,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'adjp_neglog10',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### ashr
```{r}
results_t16_t0_shrunk <- lfcShrink(dds, coef="timepoint_t16_vs_t0", type="ashr")
summary(results_t16_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t16_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t16_t0_restricted <- add_annotations_to_results(results_t16_t0_restricted)
selected_genes <- as.character(results_t16_t0_restricted$symbol)

EnhancedVolcano(results_t16_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



### P-Adjusted Analysis
```{r}
resSig <- subset(results_t16_t0, padj < 0.1) 
summary(resSig)
```

### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        



## Time -- ZM T0 to T20

### Unrestricted
#### Summary
```{r}
results_t20_t0 <- results(dds, contrast = c("timepoint", "t20", "t0"))

results_t20_t0 = add_annotations_to_results(results_t20_t0)

summary(results_t20_t0)

```

### Restricted
#### Summary
```{r}
results_t20_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t20", "t0")
)
summary(results_t20_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t20_t0_restricted)
```

        

#### V- Plot

```{r}
results_t20_t0_restricted <- add_annotations_to_results(results_t20_t0_restricted)
selected_genes <- as.character(results_t20_t0_restricted$symbol)

EnhancedVolcano(results_t20_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t20_t0_shrunk <- lfcShrink(dds, coef="timepoint_t20_vs_t0", type="apeglm")
summary(results_t20_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t20_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t20_t0_restricted <- add_annotations_to_results(results_t20_t0_restricted)
selected_genes <- as.character(results_t20_t0_restricted$symbol)

EnhancedVolcano(results_t20_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### ashr
```{r}
results_t20_t0_shrunk <- lfcShrink(dds, coef="timepoint_t20_vs_t0", type="ashr")
summary(results_t20_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t20_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t20_t0_restricted <- add_annotations_to_results(results_t20_t0_restricted)
selected_genes <- as.character(results_t20_t0_restricted$symbol)

EnhancedVolcano(results_t20_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



### P-Adjusted Analysis
```{r}
resSig <- subset(results_t20_t0, padj < 0.1) 
summary(resSig)
```

### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        



## Time -- ZM T0 to T24

### Unrestricted
#### Summary
```{r}
results_t24_t0 <- results(dds, contrast = c("timepoint", "t24", "t0"))

results_t24_t0 = add_annotations_to_results(results_t24_t0)

summary(results_t24_t0)

```

### Restricted
#### Summary
```{r}
results_t24_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t24", "t0")
)
summary(results_t24_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t24_t0_restricted)
```

        

#### V- Plot

```{r}
results_t24_t0_restricted <- add_annotations_to_results(results_t24_t0_restricted)
selected_genes <- as.character(results_t24_t0_restricted$symbol)

EnhancedVolcano(results_t24_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t24_t0_shrunk <- lfcShrink(dds, coef="timepoint_t24_vs_t0", type="apeglm")
summary(results_t24_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t24_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t24_t0_restricted <- add_annotations_to_results(results_t24_t0_restricted)
selected_genes <- as.character(results_t24_t0_restricted$symbol)

EnhancedVolcano(results_t24_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### ashr
```{r}
results_t24_t0_shrunk <- lfcShrink(dds, coef="timepoint_t24_vs_t0", type="ashr")
summary(results_t24_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t24_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t24_t0_restricted <- add_annotations_to_results(results_t24_t0_restricted)
selected_genes <- as.character(results_t24_t0_restricted$symbol)

EnhancedVolcano(results_t24_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



### P-Adjusted Analysis
```{r}
resSig <- subset(results_t24_t0, padj < 0.1) 
summary(resSig)
```

### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        



## Time -- ZM T0 to T36

### Unrestricted
#### Summary
```{r}
results_t36_t0 <- results(dds, contrast = c("timepoint", "t36", "t0"))

results_t36_t0 = add_annotations_to_results(results_t36_t0)

summary(results_t36_t0)

```

### Restricted
#### Summary
```{r}
results_t36_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t36", "t0")
)
summary(results_t36_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t36_t0_restricted)
```

        

#### V- Plot

```{r}
results_t36_t0_restricted <- add_annotations_to_results(results_t36_t0_restricted)
selected_genes <- as.character(results_t36_t0_restricted$symbol)

EnhancedVolcano(results_t36_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t36_t0_shrunk <- lfcShrink(dds, coef="timepoint_t36_vs_t0", type="apeglm")
summary(results_t36_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t36_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t36_t0_restricted <- add_annotations_to_results(results_t36_t0_restricted)
selected_genes <- as.character(results_t36_t0_restricted$symbol)

EnhancedVolcano(results_t36_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### ashr
```{r}
results_t36_t0_shrunk <- lfcShrink(dds, coef="timepoint_t36_vs_t0", type="ashr")
summary(results_t36_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t36_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t36_t0_restricted <- add_annotations_to_results(results_t36_t0_restricted)
selected_genes <- as.character(results_t36_t0_restricted$symbol)

EnhancedVolcano(results_t36_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



### P-Adjusted Analysis
```{r}
resSig <- subset(results_t36_t0, padj < 0.1) 
summary(resSig)
```

### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        



## Time -- ZM T0 to T48

### Unrestricted
#### Summary
```{r}
results_t48_t0 <- results(dds, contrast = c("timepoint", "t48", "t0"))

results_t48_t0 = add_annotations_to_results(results_t48_t0)

summary(results_t48_t0)

```

### Restricted
#### Summary
```{r}
results_t48_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t48", "t0")
)
summary(results_t48_t0_restricted)

```

#### MA Plot
```{r}

plotMA(results_t48_t0_restricted)
```

        

#### V- Plot

```{r}
results_t48_t0_restricted <- add_annotations_to_results(results_t48_t0_restricted)
selected_genes <- as.character(results_t48_t0_restricted$symbol)

EnhancedVolcano(results_t48_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


### With LFC-Shrink
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
#### APE GLM
```{r}

results_t48_t0_shrunk <- lfcShrink(dds, coef="timepoint_t48_vs_t0", type="apeglm")
summary(results_t48_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t48_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t48_t0_restricted <- add_annotations_to_results(results_t48_t0_restricted)
selected_genes <- as.character(results_t48_t0_restricted$symbol)

EnhancedVolcano(results_t48_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### ashr
```{r}
results_t48_t0_shrunk <- lfcShrink(dds, coef="timepoint_t48_vs_t0", type="ashr")
summary(results_t48_t0_shrunk)

```


#### MA Plot
```{r}

plotMA(results_t48_t0_shrunk)
```

        

#### V- Plot

```{r}
results_t48_t0_restricted <- add_annotations_to_results(results_t48_t0_restricted)
selected_genes <- as.character(results_t48_t0_restricted$symbol)

EnhancedVolcano(results_t48_t0_restricted,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                title = 'Differential expression',
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



### P-Adjusted Analysis
```{r}
resSig <- subset(results_t48_t0, padj < 0.1) 
summary(resSig)
```

### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        
        

