

---
title: "Salmon Analysis - ZM_r1to6 "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library("apeglm")
library("ashr")
library(EnhancedVolcano)
library(org.Hs.eg.db)
library("pheatmap")
library("RColorBrewer")
library("PoiClaClu")
library("limma")
getwd()
load("../../../data/ZM_r1to6_data.RData")
dds
results <- results(dds)
resultsNames(dds)

add_annotations_to_results <- function(res) {
ens.str <- substr(rownames(res), 1, 15)
res$symbol <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="SYMBOL",
                    keytype="ENSEMBL",
                    multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="ENTREZID",
                    keytype="ENSEMBL",
                    multiVals="first")
return (res)
}
```
        
# ZM Analysis {.tabset}
## Exploratory Analysis


### PCA
#### VST - Variance Stablised Transformation
```{r}
vsd_ZM <- vst(dds, blind=FALSE)
plotPCA(vsd_ZM, intgroup=c('batch', 'timepoint'))
```


#### rLog

```{r}
rld_ZM <- rlog(dds, blind=FALSE)
plotPCA(rld_ZM, intgroup=c('batch', 'timepoint'))
```


### Heatmap

#### Euclidian Distance



```{r}
sampleDists_ZM <- dist(t(assay(vsd_ZM)))

sampleDistMatrix_ZM <- as.matrix( sampleDists_ZM )
rownames(sampleDistMatrix_ZM) <- paste( vsd_ZM$timepoint, vsd_ZM$batch, sep = "_" )
colnames(sampleDistMatrix_ZM) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix_ZM,
         clustering_distance_rows = sampleDists_ZM,
         clustering_distance_cols = sampleDists_ZM,
         col = colors)
```


#### Poission Distance

```{r}
poisd_ZM <- PoissonDistance(t(counts(dds)))

samplePoisDistMatrix_ZM <- as.matrix( poisd_ZM$dd )
rownames(samplePoisDistMatrix_ZM) <- paste( vsd_ZM$timepoint, vsd_ZM$batch, sep = "_" )
colnames(samplePoisDistMatrix_ZM) <- NULL
pheatmap(samplePoisDistMatrix_ZM,
         clustering_distance_rows = poisd_ZM$dd,
         clustering_distance_cols = poisd_ZM$dd,
         col = colors)
```




        
We can use the package limma to account for the batch effect:
https://bioconductor.org/packages/release/bioc/html/limma.html

We can see that there is a clear batch effect.
```{r}
vsd_ZM <- vst(dds, blind=FALSE)
plotPCA(vsd_ZM, intgroup=c('batch', 'timepoint'))

```

Using limma we can normalise the counts for this.
```{r}
mat_ZM <- assay(vsd_ZM)
mm_ZM <- model.matrix(~timepoint, colData(vsd_ZM))
mat_ZM <- limma::removeBatchEffect(mat_ZM, batch=vsd_ZM$batch, design=mm_ZM)
assay(vsd_ZM) <- mat_ZM
plotPCA(vsd_ZM, intgroup=c('batch', 'timepoint'))

```
    




            
## Time -- ZM T0 to T16 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_t16_t0 <- results(dds, contrast = c("timepoint", "t16", "t0"))

results_t16_t0 = add_annotations_to_results(results_t16_t0)

summary(results_t16_t0)

```
##### Restricted Values
```{r}
results_t16_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t16", "t0")
)
summary(results_t16_t0_restricted)

```

MA Plot
```{r}

plotMA(results_t16_t0_restricted)
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_t16_t0_shrunk_apeglm <- lfcShrink(dds, coef="timepoint_t16_vs_t0", type="apeglm")
summary(results_t16_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t16_t0_shrunk_apeglm)
```

        

###### V-Plot

V- Plot

```{r}
results_t16_t0_shrunk_apeglm <- add_annotations_to_results(results_t16_t0_shrunk_apeglm)
selected_genes <- as.character(results_t16_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_t16_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t16_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_t16_t0_shrunk_ashr <- lfcShrink(dds, coef="timepoint_t16_vs_t0", type="ashr")
summary(results_t16_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t16_t0_shrunk_ashr)
```

        

###### V-Plot

V- Plot

```{r}
results_t16_t0_shrunk_ashr <- add_annotations_to_results(results_t16_t0_shrunk_ashr)
selected_genes <- as.character(results_t16_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_t16_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t16_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_t16_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_results_t0_t16 <- model.matrix(~ timepoint, colData(vsd_ZM))
fit_results_t0_t16 <- lmFit(mat_ZM, design_results_t0_t16)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t16
contrast.matrix <- makeContrasts(timepointt16, levels=design_results_t0_t16)
fit2_results_t0_t16 <- contrasts.fit(fit_results_t0_t16, contrast.matrix)
fit2_results_t0_t16 <- eBayes(fit2_results_t0_t16)

results_results_t0_t16 <- topTable(fit2_results_t0_t16, adjust.method="BH", number=Inf)
results_results_t0_t16 <- add_annotations_to_results(results_results_t0_t16)

results_clean__results_t0_t16 <- results_results_t0_t16[!is.na(results_results_t0_t16$logFC), ]

results_ordered_results_t0_t16 <- results_clean__results_t0_t16[order(results_clean__results_t0_t16$logFC), ]
```

```{r}
selected_genes_results_t0_t16 <- as.character(results_ordered_results_t0_t16$symbol)

EnhancedVolcano(
    results_ordered_results_t0_t16,
    lab = selected_genes_results_t0_t16,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t16 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_results_t0_t16 <- add_annotations_to_results(results_ordered_results_t0_t16)
results_ordered_results_t0_t16_df <- as.data.frame(results_ordered_results_t0_t16)
write.csv(results_ordered_results_t0_t16_df, file = "../../../../results/batch_corrected_results_t0_t16_data.csv")

```





            
## Time -- ZM T0 to T20 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_t20_t0 <- results(dds, contrast = c("timepoint", "t20", "t0"))

results_t20_t0 = add_annotations_to_results(results_t20_t0)

summary(results_t20_t0)

```
##### Restricted Values
```{r}
results_t20_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t20", "t0")
)
summary(results_t20_t0_restricted)

```

MA Plot
```{r}

plotMA(results_t20_t0_restricted)
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_t20_t0_shrunk_apeglm <- lfcShrink(dds, coef="timepoint_t20_vs_t0", type="apeglm")
summary(results_t20_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t20_t0_shrunk_apeglm)
```

        

###### V-Plot

V- Plot

```{r}
results_t20_t0_shrunk_apeglm <- add_annotations_to_results(results_t20_t0_shrunk_apeglm)
selected_genes <- as.character(results_t20_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_t20_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t20_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_t20_t0_shrunk_ashr <- lfcShrink(dds, coef="timepoint_t20_vs_t0", type="ashr")
summary(results_t20_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t20_t0_shrunk_ashr)
```

        

###### V-Plot

V- Plot

```{r}
results_t20_t0_shrunk_ashr <- add_annotations_to_results(results_t20_t0_shrunk_ashr)
selected_genes <- as.character(results_t20_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_t20_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t20_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_t20_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_results_t0_t20 <- model.matrix(~ timepoint, colData(vsd_ZM))
fit_results_t0_t20 <- lmFit(mat_ZM, design_results_t0_t20)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t20
contrast.matrix <- makeContrasts(timepointt20, levels=design_results_t0_t20)
fit2_results_t0_t20 <- contrasts.fit(fit_results_t0_t20, contrast.matrix)
fit2_results_t0_t20 <- eBayes(fit2_results_t0_t20)

results_results_t0_t20 <- topTable(fit2_results_t0_t20, adjust.method="BH", number=Inf)
results_results_t0_t20 <- add_annotations_to_results(results_results_t0_t20)

results_clean__results_t0_t20 <- results_results_t0_t20[!is.na(results_results_t0_t20$logFC), ]

results_ordered_results_t0_t20 <- results_clean__results_t0_t20[order(results_clean__results_t0_t20$logFC), ]
```

```{r}
selected_genes_results_t0_t20 <- as.character(results_ordered_results_t0_t20$symbol)

EnhancedVolcano(
    results_ordered_results_t0_t20,
    lab = selected_genes_results_t0_t20,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t20 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_results_t0_t20 <- add_annotations_to_results(results_ordered_results_t0_t20)
results_ordered_results_t0_t20_df <- as.data.frame(results_ordered_results_t0_t20)
write.csv(results_ordered_results_t0_t20_df, file = "../../../results/batch_corrected_results_t0_t20_data.csv")

```





            
## Time -- ZM T0 to T24 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_t24_t0 <- results(dds, contrast = c("timepoint", "t24", "t0"))

results_t24_t0 = add_annotations_to_results(results_t24_t0)

summary(results_t24_t0)

```
##### Restricted Values
```{r}
results_t24_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t24", "t0")
)
summary(results_t24_t0_restricted)

```

MA Plot
```{r}

plotMA(results_t24_t0_restricted)
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_t24_t0_shrunk_apeglm <- lfcShrink(dds, coef="timepoint_t24_vs_t0", type="apeglm")
summary(results_t24_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t24_t0_shrunk_apeglm)
```

        

###### V-Plot

V- Plot

```{r}
results_t24_t0_shrunk_apeglm <- add_annotations_to_results(results_t24_t0_shrunk_apeglm)
selected_genes <- as.character(results_t24_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_t24_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t24_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_t24_t0_shrunk_ashr <- lfcShrink(dds, coef="timepoint_t24_vs_t0", type="ashr")
summary(results_t24_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t24_t0_shrunk_ashr)
```

        

###### V-Plot

V- Plot

```{r}
results_t24_t0_shrunk_ashr <- add_annotations_to_results(results_t24_t0_shrunk_ashr)
selected_genes <- as.character(results_t24_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_t24_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t24_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_t24_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_results_t0_t24 <- model.matrix(~ timepoint, colData(vsd_ZM))
fit_results_t0_t24 <- lmFit(mat_ZM, design_results_t0_t24)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t24
contrast.matrix <- makeContrasts(timepointt24, levels=design_results_t0_t24)
fit2_results_t0_t24 <- contrasts.fit(fit_results_t0_t24, contrast.matrix)
fit2_results_t0_t24 <- eBayes(fit2_results_t0_t24)

results_results_t0_t24 <- topTable(fit2_results_t0_t24, adjust.method="BH", number=Inf)
results_results_t0_t24 <- add_annotations_to_results(results_results_t0_t24)

results_clean__results_t0_t24 <- results_results_t0_t24[!is.na(results_results_t0_t24$logFC), ]

results_ordered_results_t0_t24 <- results_clean__results_t0_t24[order(results_clean__results_t0_t24$logFC), ]
```

```{r}
selected_genes_results_t0_t24 <- as.character(results_ordered_results_t0_t24$symbol)

EnhancedVolcano(
    results_ordered_results_t0_t24,
    lab = selected_genes_results_t0_t24,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t24 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_results_t0_t24 <- add_annotations_to_results(results_ordered_results_t0_t24)
results_ordered_results_t0_t24_df <- as.data.frame(results_ordered_results_t0_t24)
write.csv(results_ordered_results_t0_t24_df, file = "../../../results/batch_corrected_results_t0_t24_data.csv")

```





            
## Time -- ZM T0 to T36 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_t36_t0 <- results(dds, contrast = c("timepoint", "t36", "t0"))

results_t36_t0 = add_annotations_to_results(results_t36_t0)

summary(results_t36_t0)

```
##### Restricted Values
```{r}
results_t36_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t36", "t0")
)
summary(results_t36_t0_restricted)

```

MA Plot
```{r}

plotMA(results_t36_t0_restricted)
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_t36_t0_shrunk_apeglm <- lfcShrink(dds, coef="timepoint_t36_vs_t0", type="apeglm")
summary(results_t36_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t36_t0_shrunk_apeglm)
```

        

###### V-Plot

V- Plot

```{r}
results_t36_t0_shrunk_apeglm <- add_annotations_to_results(results_t36_t0_shrunk_apeglm)
selected_genes <- as.character(results_t36_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_t36_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t36_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_t36_t0_shrunk_ashr <- lfcShrink(dds, coef="timepoint_t36_vs_t0", type="ashr")
summary(results_t36_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t36_t0_shrunk_ashr)
```

        

###### V-Plot

V- Plot

```{r}
results_t36_t0_shrunk_ashr <- add_annotations_to_results(results_t36_t0_shrunk_ashr)
selected_genes <- as.character(results_t36_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_t36_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t36_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_t36_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_results_t0_t36 <- model.matrix(~ timepoint, colData(vsd_ZM))
fit_results_t0_t36 <- lmFit(mat_ZM, design_results_t0_t36)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t36
contrast.matrix <- makeContrasts(timepointt36, levels=design_results_t0_t36)
fit2_results_t0_t36 <- contrasts.fit(fit_results_t0_t36, contrast.matrix)
fit2_results_t0_t36 <- eBayes(fit2_results_t0_t36)

results_results_t0_t36 <- topTable(fit2_results_t0_t36, adjust.method="BH", number=Inf)
results_results_t0_t36 <- add_annotations_to_results(results_results_t0_t36)

results_clean__results_t0_t36 <- results_results_t0_t36[!is.na(results_results_t0_t36$logFC), ]

results_ordered_results_t0_t36 <- results_clean__results_t0_t36[order(results_clean__results_t0_t36$logFC), ]
```

```{r}
selected_genes_results_t0_t36 <- as.character(results_ordered_results_t0_t36$symbol)

EnhancedVolcano(
    results_ordered_results_t0_t36,
    lab = selected_genes_results_t0_t36,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t36 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_results_t0_t36 <- add_annotations_to_results(results_ordered_results_t0_t36)
results_ordered_results_t0_t36_df <- as.data.frame(results_ordered_results_t0_t36)
write.csv(results_ordered_results_t0_t36_df, file = "../../../results/batch_corrected_results_t0_t36_data.csv")

```





            
## Time -- ZM T0 to T48 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_t48_t0 <- results(dds, contrast = c("timepoint", "t48", "t0"))

results_t48_t0 = add_annotations_to_results(results_t48_t0)

summary(results_t48_t0)

```
##### Restricted Values
```{r}
results_t48_t0_restricted <-
results(
    dds,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t48", "t0")
)
summary(results_t48_t0_restricted)

```

MA Plot
```{r}

plotMA(results_t48_t0_restricted)
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_t48_t0_shrunk_apeglm <- lfcShrink(dds, coef="timepoint_t48_vs_t0", type="apeglm")
summary(results_t48_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t48_t0_shrunk_apeglm)
```

        

###### V-Plot

V- Plot

```{r}
results_t48_t0_shrunk_apeglm <- add_annotations_to_results(results_t48_t0_shrunk_apeglm)
selected_genes <- as.character(results_t48_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_t48_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t48_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_t48_t0_shrunk_ashr <- lfcShrink(dds, coef="timepoint_t48_vs_t0", type="ashr")
summary(results_t48_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

plotMA(results_t48_t0_shrunk_ashr)
```

        

###### V-Plot

V- Plot

```{r}
results_t48_t0_shrunk_ashr <- add_annotations_to_results(results_t48_t0_shrunk_ashr)
selected_genes <- as.character(results_t48_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_t48_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_t48_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_t48_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_results_t0_t48 <- model.matrix(~ timepoint, colData(vsd_ZM))
fit_results_t0_t48 <- lmFit(mat_ZM, design_results_t0_t48)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t48
contrast.matrix <- makeContrasts(timepointt48, levels=design_results_t0_t48)
fit2_results_t0_t48 <- contrasts.fit(fit_results_t0_t48, contrast.matrix)
fit2_results_t0_t48 <- eBayes(fit2_results_t0_t48)

results_results_t0_t48 <- topTable(fit2_results_t0_t48, adjust.method="BH", number=Inf)
results_results_t0_t48 <- add_annotations_to_results(results_results_t0_t48)

results_clean__results_t0_t48 <- results_results_t0_t48[!is.na(results_results_t0_t48$logFC), ]

results_ordered_results_t0_t48 <- results_clean__results_t0_t48[order(results_clean__results_t0_t48$logFC), ]
```

```{r}
selected_genes_results_t0_t48 <- as.character(results_ordered_results_t0_t48$symbol)

EnhancedVolcano(
    results_ordered_results_t0_t48,
    lab = selected_genes_results_t0_t48,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t48 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_results_t0_t48 <- add_annotations_to_results(results_ordered_results_t0_t48)
results_ordered_results_t0_t48_df <- as.data.frame(results_ordered_results_t0_t48)
write.csv(results_ordered_results_t0_t48_df, file = "../../../results/batch_corrected_results_t0_t48_data.csv")

```



