

---
title: "STAR Analysis - Noc_r1to6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/bsilke/bs-villungerlab")
library(DESeq2)
library("apeglm")
library("ashr")
library(EnhancedVolcano)
library(org.Hs.eg.db)
library("pheatmap")
library("RColorBrewer")
library("PoiClaClu")
library("limma")
library("glue")
library("Rsubread")
library("stringr")
library("DESeq2")
library(dplyr)
library(openxlsx)
load('~/bs-villungerlab/results/output_encode_1to6/Noc_star_data.RData')

dds_Noc_r1to6 <- ddseq_Noc
results <- results(dds_Noc_r1to6)
resultsNames(dds_Noc_r1to6)


add_annotations_to_results <- function(res) {
ens.str <- substr(rownames(res), 1, 15)
res$symbol <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="SYMBOL",
                    keytype="ENSEMBL",
                    multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="ENTREZID",
                    keytype="ENSEMBL",
                    multiVals="first")
return (res)
}

```
        
# Noc Analysis {.tabset}
## Exploratory Analysis


### PCA
#### VST - Variance Stablised Transformation
```{r}
vsd_Noc <- vst(dds_Noc_r1to6, blind=FALSE)
plot <- plotPCA(vsd_Noc, intgroup=c('batch', 'timepoint'))
plot + ggtitle("PCA plot highlighting batch differences for VST for Noc: r1to6")
```


#### rLog

```{r}
rld_Noc <- rlog(dds_Noc_r1to6, blind=FALSE)
plotPCA(rld_Noc, intgroup=c('batch', 'timepoint'))
```


### Heatmap

#### Euclidian Distance



```{r}
sampleDists_Noc <- dist(t(assay(vsd_Noc)))

sampleDistMatrix_Noc <- as.matrix( sampleDists_Noc )
rownames(sampleDistMatrix_Noc) <- paste( vsd_Noc$timepoint, vsd_Noc$batch, sep = "_" )
colnames(sampleDistMatrix_Noc) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix_Noc,
         clustering_distance_rows = sampleDists_Noc,
         clustering_distance_cols = sampleDists_Noc,
         col = colors)
```


#### Poission Distance

```{r}
poisd_Noc <- PoissonDistance(t(counts(dds_Noc_r1to6)))

samplePoisDistMatrix_Noc <- as.matrix( poisd_Noc$dd )
rownames(samplePoisDistMatrix_Noc) <- paste( vsd_Noc$timepoint, vsd_Noc$batch, sep = "_" )
colnames(samplePoisDistMatrix_Noc) <- NULL
pheatmap(samplePoisDistMatrix_Noc,
         clustering_distance_rows = poisd_Noc$dd,
         clustering_distance_cols = poisd_Noc$dd,
         col = colors)
```



### Batch Correction using _Limma_.
        
We can use the package _limma_ to account for the batch effect:
https://bioconductor.org/packages/release/bioc/html/limma.html

```{r}
vsd_Noc <- vst(dds_Noc_r1to6, blind=FALSE)
plotPCA(vsd_Noc, intgroup=c('batch', 'timepoint'))
```

```{r}
plot <- plotPCA(vsd_Noc, intgroup=c('batch'))
plot + ggtitle("PCA plot highlighting batch differences for VST for Noc: r1to6")
```


Using limma we can normalise the counts for this.
```{r}
mat_Noc <- assay(vsd_Noc)
mm_Noc <- model.matrix(~timepoint, colData(vsd_Noc))
mat_Noc <- limma::removeBatchEffect(mat_Noc, batch=vsd_Noc$batch, design=mm_Noc)
assay(vsd_Noc) <- mat_Noc
plotPCA(vsd_Noc, intgroup=c('batch', 'timepoint'))

```

It is easier to view the change in timepoints here.
```{r}
plotPCA(vsd_Noc, intgroup=c('timepoint'))
```
    




            
## Time -- Noc T0 to T16 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Noc_t16_t0 <- results(dds_Noc_r1to6, contrast = c("timepoint", "t16", "t0"))

results_Noc_t16_t0 = add_annotations_to_results(results_Noc_t16_t0)

summary(results_Noc_t16_t0)

```
##### Restricted Values
```{r}
results_Noc_t16_t0_restricted <-
results(
    dds_Noc_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t16", "t0")
)
summary(results_Noc_t16_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t16_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Noc_t16_t0_shrunk_apeglm <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t16_vs_t0", type="apeglm")
summary(results_Noc_t16_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t16_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t16_t0_shrunk_apeglm <- add_annotations_to_results(results_Noc_t16_t0_shrunk_apeglm)
selected_genes <- as.character(results_Noc_t16_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Noc_t16_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t16_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Noc_t16_t0_shrunk_ashr <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t16_vs_t0", type="ashr")
summary(results_Noc_t16_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t16_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t16_t0_shrunk_ashr <- add_annotations_to_results(results_Noc_t16_t0_shrunk_ashr)
selected_genes <- as.character(results_Noc_t16_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Noc_t16_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t16_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Noc_t16_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Noc_16 <- model.matrix(~ timepoint, colData(vsd_Noc))
fit_Noc_16 <- lmFit(mat_Noc, design_Noc_16)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t16
contrast.matrix <- makeContrasts(timepointt16, levels=design_Noc_16)
fit2_Noc_16 <- contrasts.fit(fit_Noc_16, contrast.matrix)
fit2_Noc_16 <- eBayes(fit2_Noc_16)

results_Noc_16 <- topTable(fit2_Noc_16, adjust.method="BH", number=Inf)
results_Noc_16 <- add_annotations_to_results(results_Noc_16)

results_clean_Noc_16 <- results_Noc_16[!is.na(results_Noc_16$logFC), ]

results_ordered_Noc_16 <- results_clean_Noc_16[order(results_clean_Noc_16$logFC), ]
```

```{r}
selected_genes_Noc_16 <- as.character(results_ordered_Noc_16$symbol)

EnhancedVolcano(
    results_ordered_Noc_16,
    lab = selected_genes_Noc_16,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled Noc t0_t16 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
# results_ordered_Noc_16 <- add_annotations_to_results(results_ordered_Noc_16)
# results_ordered_Noc_16_df <- as.data.frame(results_ordered_Noc_16)
# write.csv(results_ordered_Noc_16_df, file = "~/bs-villunger/results/batch_corrected_Noc_16_data.csv")
```





            
## Time -- Noc T0 to T20 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Noc_t20_t0 <- results(dds_Noc_r1to6, contrast = c("timepoint", "t20", "t0"))

results_Noc_t20_t0 = add_annotations_to_results(results_Noc_t20_t0)

summary(results_Noc_t20_t0)

```
##### Restricted Values
```{r}
results_Noc_t20_t0_restricted <-
results(
    dds_Noc_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t20", "t0")
)
summary(results_Noc_t20_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t20_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Noc_t20_t0_shrunk_apeglm <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t20_vs_t0", type="apeglm")
summary(results_Noc_t20_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t20_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t20_t0_shrunk_apeglm <- add_annotations_to_results(results_Noc_t20_t0_shrunk_apeglm)
selected_genes <- as.character(results_Noc_t20_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Noc_t20_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t20_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Noc_t20_t0_shrunk_ashr <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t20_vs_t0", type="ashr")
summary(results_Noc_t20_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t20_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t20_t0_shrunk_ashr <- add_annotations_to_results(results_Noc_t20_t0_shrunk_ashr)
selected_genes <- as.character(results_Noc_t20_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Noc_t20_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t20_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Noc_t20_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Noc_20 <- model.matrix(~ timepoint, colData(vsd_Noc))
fit_Noc_20 <- lmFit(mat_Noc, design_Noc_20)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t20
contrast.matrix <- makeContrasts(timepointt20, levels=design_Noc_20)
fit2_Noc_20 <- contrasts.fit(fit_Noc_20, contrast.matrix)
fit2_Noc_20 <- eBayes(fit2_Noc_20)

results_Noc_20 <- topTable(fit2_Noc_20, adjust.method="BH", number=Inf)
results_Noc_20 <- add_annotations_to_results(results_Noc_20)

results_clean_Noc_20 <- results_Noc_20[!is.na(results_Noc_20$logFC), ]

results_ordered_Noc_20 <- results_clean_Noc_20[order(results_clean_Noc_20$logFC), ]
```

```{r}
selected_genes_Noc_20 <- as.character(results_ordered_Noc_20$symbol)

EnhancedVolcano(
    results_ordered_Noc_20,
    lab = selected_genes_Noc_20,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled Noc t0_t20 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
# results_ordered_Noc_20 <- add_annotations_to_results(results_ordered_Noc_20)
# results_ordered_Noc_20_df <- as.data.frame(results_ordered_Noc_20)
# write.csv(results_ordered_Noc_20_df, file = "~/bs-villunger/results/batch_corrected_Noc_20_data.csv")
```





            
## Time -- Noc T0 to T24 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Noc_t24_t0 <- results(dds_Noc_r1to6, contrast = c("timepoint", "t24", "t0"))

results_Noc_t24_t0 = add_annotations_to_results(results_Noc_t24_t0)

summary(results_Noc_t24_t0)

```
##### Restricted Values
```{r}
results_Noc_t24_t0_restricted <-
results(
    dds_Noc_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t24", "t0")
)
summary(results_Noc_t24_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t24_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Noc_t24_t0_shrunk_apeglm <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t24_vs_t0", type="apeglm")
summary(results_Noc_t24_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t24_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t24_t0_shrunk_apeglm <- add_annotations_to_results(results_Noc_t24_t0_shrunk_apeglm)
selected_genes <- as.character(results_Noc_t24_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Noc_t24_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t24_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Noc_t24_t0_shrunk_ashr <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t24_vs_t0", type="ashr")
summary(results_Noc_t24_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t24_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t24_t0_shrunk_ashr <- add_annotations_to_results(results_Noc_t24_t0_shrunk_ashr)
selected_genes <- as.character(results_Noc_t24_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Noc_t24_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t24_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Noc_t24_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Noc_24 <- model.matrix(~ timepoint, colData(vsd_Noc))
fit_Noc_24 <- lmFit(mat_Noc, design_Noc_24)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t24
contrast.matrix <- makeContrasts(timepointt24, levels=design_Noc_24)
fit2_Noc_24 <- contrasts.fit(fit_Noc_24, contrast.matrix)
fit2_Noc_24 <- eBayes(fit2_Noc_24)

results_Noc_24 <- topTable(fit2_Noc_24, adjust.method="BH", number=Inf)
results_Noc_24 <- add_annotations_to_results(results_Noc_24)

results_clean_Noc_24 <- results_Noc_24[!is.na(results_Noc_24$logFC), ]

results_ordered_Noc_24 <- results_clean_Noc_24[order(results_clean_Noc_24$logFC), ]
```

```{r}
selected_genes_Noc_24 <- as.character(results_ordered_Noc_24$symbol)

EnhancedVolcano(
    results_ordered_Noc_24,
    lab = selected_genes_Noc_24,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled Noc t0_t24 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
# results_ordered_Noc_24 <- add_annotations_to_results(results_ordered_Noc_24)
# results_ordered_Noc_24_df <- as.data.frame(results_ordered_Noc_24)
# write.csv(results_ordered_Noc_24_df, file = "~/bs-villunger/results/batch_corrected_Noc_24_data.csv")
```





            
## Time -- Noc T0 to T36 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Noc_t36_t0 <- results(dds_Noc_r1to6, contrast = c("timepoint", "t36", "t0"))

results_Noc_t36_t0 = add_annotations_to_results(results_Noc_t36_t0)

summary(results_Noc_t36_t0)

```
##### Restricted Values
```{r}
results_Noc_t36_t0_restricted <-
results(
    dds_Noc_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t36", "t0")
)
summary(results_Noc_t36_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t36_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Noc_t36_t0_shrunk_apeglm <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t36_vs_t0", type="apeglm")
summary(results_Noc_t36_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t36_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t36_t0_shrunk_apeglm <- add_annotations_to_results(results_Noc_t36_t0_shrunk_apeglm)
selected_genes <- as.character(results_Noc_t36_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Noc_t36_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t36_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Noc_t36_t0_shrunk_ashr <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t36_vs_t0", type="ashr")
summary(results_Noc_t36_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t36_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t36_t0_shrunk_ashr <- add_annotations_to_results(results_Noc_t36_t0_shrunk_ashr)
selected_genes <- as.character(results_Noc_t36_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Noc_t36_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t36_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Noc_t36_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Noc_36 <- model.matrix(~ timepoint, colData(vsd_Noc))
fit_Noc_36 <- lmFit(mat_Noc, design_Noc_36)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t36
contrast.matrix <- makeContrasts(timepointt36, levels=design_Noc_36)
fit2_Noc_36 <- contrasts.fit(fit_Noc_36, contrast.matrix)
fit2_Noc_36 <- eBayes(fit2_Noc_36)

results_Noc_36 <- topTable(fit2_Noc_36, adjust.method="BH", number=Inf)
results_Noc_36 <- add_annotations_to_results(results_Noc_36)

results_clean_Noc_36 <- results_Noc_36[!is.na(results_Noc_36$logFC), ]

results_ordered_Noc_36 <- results_clean_Noc_36[order(results_clean_Noc_36$logFC), ]
```

```{r}
selected_genes_Noc_36 <- as.character(results_ordered_Noc_36$symbol)

EnhancedVolcano(
    results_ordered_Noc_36,
    lab = selected_genes_Noc_36,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled Noc t0_t36 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
# results_ordered_Noc_36 <- add_annotations_to_results(results_ordered_Noc_36)
# results_ordered_Noc_36_df <- as.data.frame(results_ordered_Noc_36)
# write.csv(results_ordered_Noc_36_df, file = "~/bs-villunger/results/batch_corrected_Noc_36_data.csv")
```





            
## Time -- Noc T0 to T48 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Noc_t48_t0 <- results(dds_Noc_r1to6, contrast = c("timepoint", "t48", "t0"))

results_Noc_t48_t0 = add_annotations_to_results(results_Noc_t48_t0)

summary(results_Noc_t48_t0)

```
##### Restricted Values
```{r}
results_Noc_t48_t0_restricted <-
results(
    dds_Noc_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t48", "t0")
)
summary(results_Noc_t48_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t48_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Noc_t48_t0_shrunk_apeglm <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t48_vs_t0", type="apeglm")
summary(results_Noc_t48_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t48_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t48_t0_shrunk_apeglm <- add_annotations_to_results(results_Noc_t48_t0_shrunk_apeglm)
selected_genes <- as.character(results_Noc_t48_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Noc_t48_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t48_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Noc_t48_t0_shrunk_ashr <- lfcShrink(dds_Noc_r1to6, coef="timepoint_t48_vs_t0", type="ashr")
summary(results_Noc_t48_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Noc_t48_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Noc_t48_t0_shrunk_ashr <- add_annotations_to_results(results_Noc_t48_t0_shrunk_ashr)
selected_genes <- as.character(results_Noc_t48_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Noc_t48_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Noc_t48_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Noc_t48_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Noc_48 <- model.matrix(~ timepoint, colData(vsd_Noc))
fit_Noc_48 <- lmFit(mat_Noc, design_Noc_48)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t48
contrast.matrix <- makeContrasts(timepointt48, levels=design_Noc_48)
fit2_Noc_48 <- contrasts.fit(fit_Noc_48, contrast.matrix)
fit2_Noc_48 <- eBayes(fit2_Noc_48)

results_Noc_48 <- topTable(fit2_Noc_48, adjust.method="BH", number=Inf)
results_Noc_48 <- add_annotations_to_results(results_Noc_48)

results_clean_Noc_48 <- results_Noc_48[!is.na(results_Noc_48$logFC), ]

results_ordered_Noc_48 <- results_clean_Noc_48[order(results_clean_Noc_48$logFC), ]
```

```{r}
selected_genes_Noc_48 <- as.character(results_ordered_Noc_48$symbol)

EnhancedVolcano(
    results_ordered_Noc_48,
    lab = selected_genes_Noc_48,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled Noc t0_t48 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
# results_ordered_Noc_48 <- add_annotations_to_results(results_ordered_Noc_48)
# results_ordered_Noc_48_df <- as.data.frame(results_ordered_Noc_48)
# write.csv(results_ordered_Noc_48_df, file = "~/bs-villunger/results/batch_corrected_Noc_48_data.csv")
```



