

---
title: "Salmon Analysis - Etop_r1to6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library("apeglm")
library("ashr")
library(EnhancedVolcano)
library(org.Hs.eg.db)
library("pheatmap")
library("RColorBrewer")
library("PoiClaClu")
library("limma")

load("../../../data/Etop_r1to6_data.RData")
dds_Etop_r1to6
results <- results(dds_Etop_r1to6)
resultsNames(dds_Etop_r1to6)


add_annotations_to_results <- function(res) {
ens.str <- substr(rownames(res), 1, 15)
res$symbol <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="SYMBOL",
                    keytype="ENSEMBL",
                    multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db,
                    keys=ens.str,
                    column="ENTREZID",
                    keytype="ENSEMBL",
                    multiVals="first")
return (res)
}

```
        
# Etop Analysis {.tabset}
## Exploratory Analysis


### PCA
#### VST - Variance Stablised Transformation
```{r}
vsd_Etop <- vst(dds_Etop_r1to6, blind=FALSE)
plotPCA(vsd_Etop, intgroup=c('batch', 'timepoint'))
```


#### rLog

```{r}
rld_Etop <- rlog(dds_Etop_r1to6, blind=FALSE)
plotPCA(rld_Etop, intgroup=c('batch', 'timepoint'))
```


### Heatmap

#### Euclidian Distance



```{r}
sampleDists_Etop <- dist(t(assay(vsd_Etop)))

sampleDistMatrix_Etop <- as.matrix( sampleDists_Etop )
rownames(sampleDistMatrix_Etop) <- paste( vsd_Etop$timepoint, vsd_Etop$batch, sep = "_" )
colnames(sampleDistMatrix_Etop) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix_Etop,
         clustering_distance_rows = sampleDists_Etop,
         clustering_distance_cols = sampleDists_Etop,
         col = colors)
```


#### Poission Distance

```{r}
poisd_Etop <- PoissonDistance(t(counts(dds_Etop_r1to6)))

samplePoisDistMatrix_Etop <- as.matrix( poisd_Etop$dd )
rownames(samplePoisDistMatrix_Etop) <- paste( vsd_Etop$timepoint, vsd_Etop$batch, sep = "_" )
colnames(samplePoisDistMatrix_Etop) <- NULL
pheatmap(samplePoisDistMatrix_Etop,
         clustering_distance_rows = poisd_Etop$dd,
         clustering_distance_cols = poisd_Etop$dd,
         col = colors)
```




        
We can use the package limma to account for the batch effect:
https://bioconductor.org/packages/release/bioc/html/limma.html

We can see that there is a clear batch effect.
```{r}
vsd_Etop <- vst(dds_Etop_r1to6, blind=FALSE)
plotPCA(vsd_Etop, intgroup=c('batch', 'timepoint'))

```

Using limma we can normalise the counts for this.
```{r}
mat_Etop <- assay(vsd_Etop)
mm_Etop <- model.matrix(~timepoint, colData(vsd_Etop))
mat_Etop <- limma::removeBatchEffect(mat_Etop, batch=vsd_Etop$batch, design=mm_Etop)
assay(vsd_Etop) <- mat_Etop
plotPCA(vsd_Etop, intgroup=c('batch', 'timepoint'))

```
    




            
## Time -- Etop T0 to T8 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Etop_t8_t0 <- results(dds_Etop_r1to6, contrast = c("timepoint", "t8", "t0"))

results_Etop_t8_t0 = add_annotations_to_results(results_Etop_t8_t0)

summary(results_Etop_t8_t0)

```
##### Restricted Values
```{r}
results_Etop_t8_t0_restricted <-
results(
    dds_Etop_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t8", "t0")
)
summary(results_Etop_t8_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t8_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Etop_t8_t0_shrunk_apeglm <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t8_vs_t0", type="apeglm")
summary(results_Etop_t8_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t8_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t8_t0_shrunk_apeglm <- add_annotations_to_results(results_Etop_t8_t0_shrunk_apeglm)
selected_genes <- as.character(results_Etop_t8_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Etop_t8_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t8_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Etop_t8_t0_shrunk_ashr <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t8_vs_t0", type="ashr")
summary(results_Etop_t8_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t8_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t8_t0_shrunk_ashr <- add_annotations_to_results(results_Etop_t8_t0_shrunk_ashr)
selected_genes <- as.character(results_Etop_t8_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Etop_t8_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t8_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Etop_t8_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Etop_8 <- model.matrix(~ timepoint, colData(vsd_Etop))
fit_Etop_8 <- lmFit(mat_Etop, design_Etop_8)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t8
contrast.matrix <- makeContrasts(timepointt8, levels=design_Etop_8)
fit2_Etop_8 <- contrasts.fit(fit_Etop_8, contrast.matrix)
fit2_Etop_8 <- eBayes(fit2_Etop_8)

results_Etop_8 <- topTable(fit2_Etop_8, adjust.method="BH", number=Inf)
results_Etop_8 <- add_annotations_to_results(results_Etop_8)

results_clean_Etop_8 <- results_Etop_8[!is.na(results_Etop_8$logFC), ]

results_ordered_Etop_8 <- results_clean_Etop_8[order(results_clean_Etop_8$logFC), ]
```

```{r}
selected_genes_Etop_8 <- as.character(results_ordered_Etop_8$symbol)

EnhancedVolcano(
    results_ordered_Etop_8,
    lab = selected_genes_Etop_8,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t8 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_Etop_8 <- add_annotations_to_results(results_ordered_Etop_8)
results_ordered_Etop_8_df <- as.data.frame(results_ordered_Etop_8)
write.csv(results_ordered_Etop_8_df, file = "../../../../results/batch_corrected_Etop_8_data.csv")
```





            
## Time -- Etop T0 to T12 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Etop_t12_t0 <- results(dds_Etop_r1to6, contrast = c("timepoint", "t12", "t0"))

results_Etop_t12_t0 = add_annotations_to_results(results_Etop_t12_t0)

summary(results_Etop_t12_t0)

```
##### Restricted Values
```{r}
results_Etop_t12_t0_restricted <-
results(
    dds_Etop_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t12", "t0")
)
summary(results_Etop_t12_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t12_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Etop_t12_t0_shrunk_apeglm <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t12_vs_t0", type="apeglm")
summary(results_Etop_t12_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t12_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t12_t0_shrunk_apeglm <- add_annotations_to_results(results_Etop_t12_t0_shrunk_apeglm)
selected_genes <- as.character(results_Etop_t12_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Etop_t12_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t12_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Etop_t12_t0_shrunk_ashr <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t12_vs_t0", type="ashr")
summary(results_Etop_t12_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t12_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t12_t0_shrunk_ashr <- add_annotations_to_results(results_Etop_t12_t0_shrunk_ashr)
selected_genes <- as.character(results_Etop_t12_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Etop_t12_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t12_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Etop_t12_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Etop_12 <- model.matrix(~ timepoint, colData(vsd_Etop))
fit_Etop_12 <- lmFit(mat_Etop, design_Etop_12)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t12
contrast.matrix <- makeContrasts(timepointt12, levels=design_Etop_12)
fit2_Etop_12 <- contrasts.fit(fit_Etop_12, contrast.matrix)
fit2_Etop_12 <- eBayes(fit2_Etop_12)

results_Etop_12 <- topTable(fit2_Etop_12, adjust.method="BH", number=Inf)
results_Etop_12 <- add_annotations_to_results(results_Etop_12)

results_clean_Etop_12 <- results_Etop_12[!is.na(results_Etop_12$logFC), ]

results_ordered_Etop_12 <- results_clean_Etop_12[order(results_clean_Etop_12$logFC), ]
```

```{r}
selected_genes_Etop_12 <- as.character(results_ordered_Etop_12$symbol)

EnhancedVolcano(
    results_ordered_Etop_12,
    lab = selected_genes_Etop_12,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t12 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_Etop_12 <- add_annotations_to_results(results_ordered_Etop_12)
results_ordered_Etop_12_df <- as.data.frame(results_ordered_Etop_12)
write.csv(results_ordered_Etop_12_df, file = "../../../../results/batch_corrected_Etop_12_data.csv")
```





            
## Time -- Etop T0 to T16 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Etop_t16_t0 <- results(dds_Etop_r1to6, contrast = c("timepoint", "t16", "t0"))

results_Etop_t16_t0 = add_annotations_to_results(results_Etop_t16_t0)

summary(results_Etop_t16_t0)

```
##### Restricted Values
```{r}
results_Etop_t16_t0_restricted <-
results(
    dds_Etop_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t16", "t0")
)
summary(results_Etop_t16_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t16_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Etop_t16_t0_shrunk_apeglm <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t16_vs_t0", type="apeglm")
summary(results_Etop_t16_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t16_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t16_t0_shrunk_apeglm <- add_annotations_to_results(results_Etop_t16_t0_shrunk_apeglm)
selected_genes <- as.character(results_Etop_t16_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Etop_t16_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t16_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Etop_t16_t0_shrunk_ashr <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t16_vs_t0", type="ashr")
summary(results_Etop_t16_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t16_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t16_t0_shrunk_ashr <- add_annotations_to_results(results_Etop_t16_t0_shrunk_ashr)
selected_genes <- as.character(results_Etop_t16_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Etop_t16_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t16_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Etop_t16_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Etop_16 <- model.matrix(~ timepoint, colData(vsd_Etop))
fit_Etop_16 <- lmFit(mat_Etop, design_Etop_16)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t16
contrast.matrix <- makeContrasts(timepointt16, levels=design_Etop_16)
fit2_Etop_16 <- contrasts.fit(fit_Etop_16, contrast.matrix)
fit2_Etop_16 <- eBayes(fit2_Etop_16)

results_Etop_16 <- topTable(fit2_Etop_16, adjust.method="BH", number=Inf)
results_Etop_16 <- add_annotations_to_results(results_Etop_16)

results_clean_Etop_16 <- results_Etop_16[!is.na(results_Etop_16$logFC), ]

results_ordered_Etop_16 <- results_clean_Etop_16[order(results_clean_Etop_16$logFC), ]
```

```{r}
selected_genes_Etop_16 <- as.character(results_ordered_Etop_16$symbol)

EnhancedVolcano(
    results_ordered_Etop_16,
    lab = selected_genes_Etop_16,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t16 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_Etop_16 <- add_annotations_to_results(results_ordered_Etop_16)
results_ordered_Etop_16_df <- as.data.frame(results_ordered_Etop_16)
write.csv(results_ordered_Etop_16_df, file = "../../../../results/batch_corrected_Etop_16_data.csv")
```





            
## Time -- Etop T0 to T24 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Etop_t24_t0 <- results(dds_Etop_r1to6, contrast = c("timepoint", "t24", "t0"))

results_Etop_t24_t0 = add_annotations_to_results(results_Etop_t24_t0)

summary(results_Etop_t24_t0)

```
##### Restricted Values
```{r}
results_Etop_t24_t0_restricted <-
results(
    dds_Etop_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t24", "t0")
)
summary(results_Etop_t24_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t24_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Etop_t24_t0_shrunk_apeglm <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t24_vs_t0", type="apeglm")
summary(results_Etop_t24_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t24_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t24_t0_shrunk_apeglm <- add_annotations_to_results(results_Etop_t24_t0_shrunk_apeglm)
selected_genes <- as.character(results_Etop_t24_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Etop_t24_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t24_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Etop_t24_t0_shrunk_ashr <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t24_vs_t0", type="ashr")
summary(results_Etop_t24_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t24_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t24_t0_shrunk_ashr <- add_annotations_to_results(results_Etop_t24_t0_shrunk_ashr)
selected_genes <- as.character(results_Etop_t24_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Etop_t24_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t24_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Etop_t24_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Etop_24 <- model.matrix(~ timepoint, colData(vsd_Etop))
fit_Etop_24 <- lmFit(mat_Etop, design_Etop_24)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t24
contrast.matrix <- makeContrasts(timepointt24, levels=design_Etop_24)
fit2_Etop_24 <- contrasts.fit(fit_Etop_24, contrast.matrix)
fit2_Etop_24 <- eBayes(fit2_Etop_24)

results_Etop_24 <- topTable(fit2_Etop_24, adjust.method="BH", number=Inf)
results_Etop_24 <- add_annotations_to_results(results_Etop_24)

results_clean_Etop_24 <- results_Etop_24[!is.na(results_Etop_24$logFC), ]

results_ordered_Etop_24 <- results_clean_Etop_24[order(results_clean_Etop_24$logFC), ]
```

```{r}
selected_genes_Etop_24 <- as.character(results_ordered_Etop_24$symbol)

EnhancedVolcano(
    results_ordered_Etop_24,
    lab = selected_genes_Etop_24,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t24 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_Etop_24 <- add_annotations_to_results(results_ordered_Etop_24)
results_ordered_Etop_24_df <- as.data.frame(results_ordered_Etop_24)
write.csv(results_ordered_Etop_24_df, file = "../../../../results/batch_corrected_Etop_24_data.csv")
```





            
## Time -- Etop T0 to T48 {.tabset}

### Exploratory Analysis {.tabset}

#### Summary
##### General Summary
```{r}
results_Etop_t48_t0 <- results(dds_Etop_r1to6, contrast = c("timepoint", "t48", "t0"))

results_Etop_t48_t0 = add_annotations_to_results(results_Etop_t48_t0)

summary(results_Etop_t48_t0)

```
##### Restricted Values
```{r}
results_Etop_t48_t0_restricted <-
results(
    dds_Etop_r1to6,
    alpha = 0.1,
    ## We are looking for genes which have at least doubled, or decreased by more than half
    lfcThreshold = 1,
    contrast = c("timepoint", "t48", "t0")
)
summary(results_Etop_t48_t0_restricted)

```

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t48_t0_restricted, ylim=c(-3,3))
```

        




### Differential Expression Analysis {.tabset}
It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.

#### APEGLM
```{r}

results_Etop_t48_t0_shrunk_apeglm <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t48_vs_t0", type="apeglm")
summary(results_Etop_t48_t0_shrunk_apeglm)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t48_t0_shrunk_apeglm, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t48_t0_shrunk_apeglm <- add_annotations_to_results(results_Etop_t48_t0_shrunk_apeglm)
selected_genes <- as.character(results_Etop_t48_t0_shrunk_apeglm$symbol)

EnhancedVolcano(results_Etop_t48_t0_shrunk_apeglm,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t48_t0_shrunk_apeglm',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```



#### ASHR
```{r}
results_Etop_t48_t0_shrunk_ashr <- lfcShrink(dds_Etop_r1to6, coef="timepoint_t48_vs_t0", type="ashr")
summary(results_Etop_t48_t0_shrunk_ashr)

```
###### MA - Plot

MA Plot
```{r}

DESeq2::plotMA(results_Etop_t48_t0_shrunk_ashr, ylim=c(-3,3))
```

        

###### V-Plot

V- Plot

```{r}
results_Etop_t48_t0_shrunk_ashr <- add_annotations_to_results(results_Etop_t48_t0_shrunk_ashr)
selected_genes <- as.character(results_Etop_t48_t0_shrunk_ashr$symbol)

EnhancedVolcano(results_Etop_t48_t0_shrunk_ashr,
                lab = selected_genes,
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-8,8),
                ylab = expression(paste('-Log'[10],' adj P')),
                title = 'Differential expression for results_Etop_t48_t0_shrunk_ashr',
                pCutoff = 0.05,
                # at least double, or less than half
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 3.0)
```


#### P-Adjusted Analysis

```{r}
resSig <- subset(results_Etop_t48_t0, padj < 0.1) 
summary(resSig)
```

###### Strongest Down Regulated Genes
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```

###### Strongest up regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```
        






### Batch Effects 

##### Differential expression analysis with limma
After we have adjusted counts for the batch effect, 
we can see if different genes are highlighted for differential expression

```{r}
design_Etop_48 <- model.matrix(~ timepoint, colData(vsd_Etop))
fit_Etop_48 <- lmFit(mat_Etop, design_Etop_48)
# Replace 'treatment' and 'control' with your specific time points.
# Demonstrates difference between t0 and t48
contrast.matrix <- makeContrasts(timepointt48, levels=design_Etop_48)
fit2_Etop_48 <- contrasts.fit(fit_Etop_48, contrast.matrix)
fit2_Etop_48 <- eBayes(fit2_Etop_48)

results_Etop_48 <- topTable(fit2_Etop_48, adjust.method="BH", number=Inf)
results_Etop_48 <- add_annotations_to_results(results_Etop_48)

results_clean_Etop_48 <- results_Etop_48[!is.na(results_Etop_48$logFC), ]

results_ordered_Etop_48 <- results_clean_Etop_48[order(results_clean_Etop_48$logFC), ]
```

```{r}
selected_genes_Etop_48 <- as.character(results_ordered_Etop_48$symbol)

EnhancedVolcano(
    results_ordered_Etop_48,
    lab = selected_genes_Etop_48,
    x = 'logFC',
    y = 'adj.P.Val',
    title = 'Batch controlled ZM t0_t48 (limma)',
    ylab = expression(paste('-Log'[10],' adj P')),       # Y-axis label
    # with adj p cutoff of 0.05
    pCutoff = 0.05,
    # at least double, or less than half
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 3.0
)

```

```{r include=FALSE}
results_ordered_Etop_48 <- add_annotations_to_results(results_ordered_Etop_48)
results_ordered_Etop_48_df <- as.data.frame(results_ordered_Etop_48)
write.csv(results_ordered_Etop_48_df, file = "../../../../results/batch_corrected_Etop_48_data.csv")
```



